Displacement of fishing effort by Large Scale Marine Protected Areas

### 2018-10-05 rmd/Manuscript.Rmd
Added some slides for ERE lunch

```{r, fig.cap = "\\label{fig:redist_trend_ps}Monthly relative allocation of fishing effort by PIPA-fishing vessels before and after PIPA for 9 EEZs, PIPA, the high seas and 'other EEZs'."}
displacement_data %>% 
  filter(gear == "purse_seines") %>% 
  ggplot(aes(x = date, y = h_prop, group = post)) +
  geom_smooth(method = "lm", color = "black", size = 0.5, linetype = "dashed") +
  geom_point(size = 1) +
  geom_vline(xintercept = lubridate::date(paste(2015, 1, 1, sep = "/")), linetype = "dashed") +
  scale_fill_brewer(palette = "Paired") +
  ggtheme_plot() +
  facet_wrap(~country, scales = "free_y")
```

>>>>

```{r, fig.cap = "\\label{fig:redist_trend_ps}Monthly relative allocation of fishing effort by PIPA-fishing vessels before and after PIPA for 9 EEZs, PIPA, the high seas and 'other EEZs'."}
displacement_data %>% 
  filter(gear == "purse_seines",
         h_prop > 0) %>%
  mutate(country = as.character(country),
         country = ifelse(country %in% c("PIPA", "KIR", "HS", "ASM", "COK", "FSM", "MHL", "NRU", "PNG", "SLB", "TKL", "TUV", "UMI"), country, "OTH"),
         country = forcats::fct_relevel(country, "OTH", after = Inf)) %>% 
  ggplot(aes(x = date, y = h_prop, group = post)) +
  geom_smooth(method = "lm", color = "black", size = 0.5, linetype = "dashed") +
  geom_point(size = 1) +
  geom_vline(xintercept = lubridate::date(paste(2015, 1, 1, sep = "/")), linetype = "dashed") +
  scale_fill_brewer(palette = "Paired") +
  ggtheme_plot() +
  facet_wrap(~country, scales = "free_y")
```

### 2018-10-05 slides/ERE_lunch.Rmd
Added some slides for ERE lunch

---
title: "Displacement of fishing effort by Large Scale Marine Protected Areas"
author:
  - Juan Carlos Villaseñor-Derbez^1^
  - John Lynham^2^
institute:
  - ^1^Bren School of Environmental Science and Management, UC Santa Barbara
  - ^2^Department of Economics, University of Hawaii at Manoa
date: "5/10/2018"
output: 
  beamer_presentation: 
    toc: yes
    slide_level: 2
    citation_package: biblatex
bibliography: references.bib
biblio-style: authoryear
---

@@

@@

source(here::here("scripts", "st_rotate.R"))
```

@@

- Marine Protected Areas (MPAs): Spatial management of fishing effort
-Areas > 250,000 Km^2^ are Large Scale (LSMPAs) \footcite{toonen_2013}
  - Implemented in the pelagic environment
  - Dominant human activity is industrial fishing \footcite{gray_2017}

```{r}
mpas <- sf::read_sf(here::here("raw_data", "spatial", "WDPA_Mar2018"),
        layer = "WDPA_Mar2018_marine-shapefile-polygons",
        quiet = T,
        stringsAsFactors = F) %>% 
  janitor::clean_names() %>% 
  filter(gis_m_area > 250000)

coast <- rnaturalearth::ne_countries(returnclass = "sf")
```

@@

@@

- Phoenix Island Protected Area
  - Part of Kiribati EEZ
  - Implemented in 2015
  - Kiribati is part of the PNA, along with other 8 countries

```{r}
eez <- sf::read_sf(dsn = here::here("data", "spatial", "PNA_EEZ"), layer = "PNA_EEZ") %>% 
  sf::st_transform(crs = "+proj=longlat +datum=WGS84 +no_defs") %>% 
  st_rotate() %>% 
  st_buffer(0) %>% 
  group_by(ISO_Ter1) %>% 
  summarize()

pipa <- sf::read_sf(dsn = here::here("data", "spatial", "PIPA"), layer = "PIPA") %>% 
  sf::st_transform(crs = "+proj=longlat +datum=WGS84 +no_defs") %>% 
  st_rotate()
```

```{r, fig.height = 4}
ggplot() +
  geom_sf(data = eez, fill = "gray", alpha = 0.5) +
  geom_sf(data = pipa, fill = "red") +
  ggtheme_plot()
```

- Phoenix Island Protected Area
  - Part of Kiribati EEZ
  - Implemented in 2015
  - Kiribati is part of the PNA, along with other 8 countries

- Recent focus of *Blue paradox* \footcite{mcdermott_2018}

```{r, fig.height = 4}
displacement_data <- readRDS(file = here::here("data", "displacement_data.rds")) %>% 
  filter(year < 2017)

displacement_data %>% 
  filter(gear == "purse_seines",
         country == "PIPA") %>% 
  ggplot(aes(x = date, y = h_prop, group = post)) +
  geom_smooth(method = "loess", color = "black", linetype = "dashed") +
  geom_point(aes(color = post), size = 2) +
  geom_vline(xintercept = lubridate::date(paste(2015, 1, 1, sep = "/")), linetype = "dashed") +
  scale_fill_brewer(palette = "Paired") +
  ggtheme_plot() +
  ylab("Proportion of total hours by vessels") +
  scale_y_continuous(labels = scales::percent) +
  scale_color_brewer(palette = "Set1")
```

```{r}
effort_by_vessel <- readRDS(file = here::here("data", "effort_by_vessel.rds")) %>% 
  filter(year < 2018)
```

# Objectives

- How do vessel-level behavior change due to PIPA implementation?
- What happens to the displaced fishing effort?

@@

@@

$$
y_{i,t} = \alpha + \beta_1 Post_t + \beta_2 Treat_i + \beta_3 Post_t \times Treat_i + \mu_1Y_t + \mu_2Y_t^2 + \phi_t + \gamma_i + \epsilon_{i,t}
$$

@@

# Preliminary results

# Future work

>>>>

---
title: "Displacement of fishing effort by Large Scale Marine Protected Areas"
author:
  - Juan Carlos Villaseñor-Derbez^1^
  - John Lynham^2^
institute:
  - ^1^Bren School of Environmental Science and Management, UC Santa Barbara
  - ^2^Department of Economics, University of Hawaii at Manoa
date: "5/10/2018"
output: 
  beamer_presentation: 
    slide_level: 2
    citation_package: biblatex
bibliography: references.bib
biblio-style: authoryear
header-includes:
  - \usepackage{booktabs}
---

@@

options(knitr.table.format = "latex")

@@

source(here::here("scripts", "st_rotate.R"))
source(here::here("scripts", "sfc_as_cols.R"))
```

@@

- Marine Protected Areas (MPAs): Spatial management of fishing effort
- Areas > 250,000 Km^2^ are Large Scale (LSMPAs) \footcite{toonen_2013}
  - Industrial fishing largest human activity in pelagic environment \footcite{gray_2017}
  - Recent widespread implementation, unknown implication for fisheries

```{r}
countries <- c("PIPA", "KIR", "HS", "ASM", "COK", "FSM", "MHL", "NRU", "PNG", "SLB", "TKL", "TUV", "UMI")

mpas <- sf::read_sf(here::here("raw_data", "spatial", "WDPA_Mar2018"),
        layer = "WDPA_Mar2018_marine-shapefile-polygons",
        quiet = T,
        stringsAsFactors = F) %>% 
  janitor::clean_names() %>% 
  filter(gis_m_area > 250000)

coast <- rnaturalearth::ne_countries(returnclass = "sf")

small_coast <- coast %>% 
  filter(sov_a3 %in% countries)

eez <- read_sf(dsn = here::here("data", "spatial", "EEZ_subset"),
               layer = "EEZ_subset") %>% 
  filter(ISO_Ter1 %in% countries) %>% 
  mutate(KIR = ISO_Ter1 == "KIR")

pipa <- sf::read_sf(dsn = here::here("data", "spatial", "PIPA"), layer = "PIPA") %>% 
  sf::st_transform(crs = "+proj=longlat +datum=WGS84 +no_defs") %>% 
  st_rotate()

effort_by_vessel <- readRDS(file = here::here("data", "effort_by_vessel.rds")) %>% 
  filter(year < 2017)

displacement_data <- readRDS(file = here::here("data", "displacement_data.rds")) %>% 
  filter(year < 2017)
```

@@

## LSMPAs

- Erroneously assumed to have little social implications due to their remoteness \footcite{agardy_2011}
- *Blue paradox* shows preemptive fishing\footcite{mcdermott_2018}:

```{r, fig.cap = "Preemtive fishing due to MPA implementation (Modified from McDermott et al (2018))", out.width = "180px"}
knitr::include_graphics(path = "img/BlueParadox.jpg")
```

## MPAs and fishing effort

Models range from *cookie-cutter* approach to spatially explicit reallocation of fishing effort based on habitat characteristics:

- All these focus on the long term equilibrium \footcite{white_2013}
- Resource users may show idiosyncratic responses \footcite{cabral_2017}
- Redistribution of fishing effort may not be optimal, especially over the first years \footcite{stevenson_2013}
- Not accounting for fisher's behavior may lead to unexpected outcomes \footcite{smith_2003}

@@

- Phoenix Island Protected Area
  - Belongs to Kiribati
  - Implemented in 2015
  - Kiribati is part of the PNA, along with other 8 countries

```{r, fig.height = 4}
ggplot() +
  geom_sf(data = small_coast) +
  geom_sf(data = eez, aes(color = KIR, linetype = !PNA), fill = "transparent") +
  geom_sf(data = pipa, fill = "red") +
  ggtheme_plot() +
  scale_color_manual(values = c("black", "red")) +
  theme(legend.position = "none")
```

# Questions

## Questions

- How does vessel-level behavior change due to PIPA implementation?
- What happens to the displaced fishing effort?

@@

- On-board Automatic Identification Systems (AIS)
- Global georeferenced vessel positions (3.1 billion and growing):
  - Activity (fishing / not fishing)
  - Time (hours)
  - Vessel characteristics (flag, gear, length, width)

```{r, fig.cap = "Sample track of Chinese longliner (10K fishing points of ~400K total)", fig.height = 4}
readRDS(file = "small_data/sample_track.rds") %>% 
  ggplot() +
  geom_sf(data = small_coast) +
  geom_sf(data = eez, aes(color = KIR, linetype = !PNA), fill = "transparent") +
  geom_sf(data = pipa, fill = "red") +
  geom_point(aes(x = lon, y = lat), shape = ".") +
  ggtheme_plot() +
  scale_color_manual(values = c("black", "red")) +
  theme(legend.position = "none")
```

Two groups:

- Treated
  - Vessels who fished inside PIPA at least once before closure
  - Continued to fish elsewhere after implementation of PIPA
- Control:
  - Vessels never fished within PIPA waters
  - Vessels belong to other PNA countries
  - Vessels have fished in surrounding areas (*i.e.* PNA-countries' EEZ) before and after PIPA closure

- Over 45 million individual AIS messages (positions)

```{r, echo = F, fig.cap = "\\label{fig:all_vessels}Fishing hours and number of vessels by month for all vessels. Vertical dashed line indicates PIPA closure."}
effort_by_vessel %>% 
  filter(baci_strict) %>%
  group_by(year, gear, treated, post, month) %>% 
  summarize(total_hours = sum(hours),
            mean_hours = mean(hours),
            n_mmsi = n()) %>% 
  ungroup() %>% 
  mutate(date = lubridate::date(paste(year, month, "01", sep = "/"))) %>% 
  select(-year) %>% 
  gather(variable, value, -c(date, month, post, gear, treated)) %>% 
  ggplot(aes(x = date, y = value, color = treated)) +
  geom_line() +
  geom_point() +
  facet_wrap(gear~variable, scales = "free", ncol = 3) +
  ggtheme_plot() +
  scale_color_brewer(palette = "Set1") +
  geom_vline(xintercept = lubridate::date(paste(2015, 01, "01", sep = "/")), linetype = "dashed")
```

@@

### Change in vessel-level behavior with a DiD

$$
y_{i,t} = \alpha + \beta_1 Post_t + \beta_2 Treat_i + \beta_3 Post_t \times Treat_i + \epsilon_{i,t}
$$

- $y_{i,t}$ monthly fishing hours by vessel $i$ in time period $t$
- $Post_t$ before-after PIPA dummy
- $Treat_i$ treatment dummy
- $\beta_3$ is our DiD estimate
- month, flag, year controls

### Spatial redistribution

- Treated vessels only

@@

- $y_{i,t}$ proportion of fishing hours that country $i$ receives at time $t$
- $Post_t$ before-after PIPA dummy
- $Country_i$ country dummy

# Preliminary results \newline (Focusing on purse seiners for now)

## Change in fishing

```{r, echo = F, fig.cap = "\\label{fig:all_vessels}Fishing hours and number of vessels by month for all vessels. Vertical dashed line indicates PIPA closure.", fig.heihgt = 3.5}

effort_by_vessel %>% 
  filter(baci_strict,
         gear == "purse_seines") %>% 
  arrange(!treated) %>% 
  ggplot(aes(x = date, y = hours, color = treated)) +
  geom_smooth(method = lm, formula = y ~ splines::bs(x, 4), aes(group = paste(treated, post))) +
  geom_point(size = 1, position = position_jitter(width = 1, height = 0), alpha = 0.5) +
  ggtheme_plot() +
  scale_color_brewer(palette = "Set1") +
  geom_vline(xintercept = lubridate::date(paste(2015, 01, "01", sep = "/")), linetype = "dashed")
```

## Change in fishing

```{r, results = "asis"}
model_data <- effort_by_vessel %>% 
  filter(baci_strict,
         gear == "purse_seines") %>% 
  mutate(post = post * 1,
         treated = treated * 1,
         year2 = year^2,
         mmsi_c = as.factor(mmsi))

m1 <- lm(hours ~ post*treated, data = model_data)
m2 <- lm(hours ~ post*treated + month_c, data = model_data)
m3 <- lm(hours ~ post*treated + month_c + year + year2, data = model_data)
m4 <- lm(hours ~ post*treated + month_c + year + year2 + flag, data = model_data)

models <- list(m1, m2, m3, m4)

stargazer::stargazer(models,
                     se = commarobust::makerobustseslist(models),
                     t.auto = T,
                     p.auto = T,
                     type = "latex",
                     omit = c("year", "flag", "month_c"),
                     add.lines = list(c("Months", "No", "Yes", "Yes", "Yes"),
                                      c("Year", "No", "No", "Yes", "Yes"),
                                      c("Country", "No", "No", "No", "Yes")),
                     omit.stat = c("adj.rsq", "f", "ser"),
                     header = F,
                     title = "\\label{tab:purse}Fishing hours from GFW for purse seiners (n = 106; 38 control, 68 treatment).",
                     column.sep.width = "1pt",
                     font.size = "tiny")
```

## Effort redistribution

```{r, fig.cap = "\\label{fig:redist_trend_ps}Monthly relative allocation of fishing effort by PIPA-vessels.", fig.height = 4}
displacement_data %>% 
  filter(gear == "purse_seines",
         h_prop > 0) %>%
  mutate(country = as.character(country),
         country = ifelse(country %in% countries, country, "OTH"),
         country = forcats::fct_relevel(country, c("PIPA", "KIR", "HS")),
         country = forcats::fct_relevel(country, "OTH", after = Inf)) %>% 
  ggplot(aes(x = date, y = h_prop, group = post)) +
  geom_smooth(method = "lm", color = "black", size = 0.5, linetype = "dashed") +
  geom_point(size = 1) +
  geom_vline(xintercept = lubridate::date(paste(2015, 1, 1, sep = "/")), linetype = "dashed") +
  scale_fill_brewer(palette = "Paired") +
  ggtheme_plot() +
  facet_wrap(~country, scales = "free_y")
```

## Displacement

```{r}
# Dont have this here. Just write the model specification

disp_mod_ps_data <- displacement_data %>% 
  filter(gear == "purse_seines",
         h_prop > 0) %>%
  mutate(country = as.character(country),
         country = ifelse(country %in% countries, country, "OTH"),
         country = forcats::fct_relevel(country, c("PIPA", "KIR", "HS")),
         country = forcats::fct_relevel(country, "OTH", after = Inf),
         post = post *1,
         year2 = year^2,
         month_c = as.character(month))

disp_mod_ps1 <- lm(h_prop ~ post * country, data = disp_mod_ps_data)

commarobust::commarobust(disp_mod_ps1) %>% 
  as.data.frame() %>% 
  mutate(term = rownames(.)) %>% 
  mutate(b = str_detect(term, "Intercept"),
         c = str_detect(term, "post"),
         a = b*c,
         p = `Pr(>|t|)`,
         p1 = ifelse(p < 0.1, "*", ""),
         p2 = ifelse(p < 0.05, "*", ""),
         p3 = ifelse(p < 0.01, "*", ""),
    h_prop = paste0(formatC(Estimate, digits = 3, format = "f"), p1, p2, p3)) %>% 
  filter(b+c > 0) %>% 
  select(term, h_prop) %>% 
  knitr::kable(booktabs = T,
               caption = "Change in the relative allocation of fishing hours by purse seiners") %>% 
  kableExtra::kable_styling(latex_options = "HOLD_position",
                            font_size = 7)
```

$R^2 = 0.544^{***}; n = 707$

## Displacement

```{r, fig.cap = "\\label{fig:map_change_ps}Spatial representation of the mean change in the monthly allocation of fishing effort.", fig.height = 4}
ba_disp_ps <- displacement_data %>% 
  filter(gear == "purse_seines") %>% 
  group_by(country, post) %>% 
  summarize(h_prop_m = mean(h_prop)) %>% 
  spread(post, h_prop_m) %>% 
  mutate(change = (`TRUE` - `FALSE`) * 100) %>% 
  select(country, change)

pipa_map_ba_disp <- pipa %>% 
  mutate(country = "PIPA",
         PNA = F) %>% 
  left_join(ba_disp_ps, by = "country") %>% 
  select(country, change, PNA)

HS_disp_poly <- matrix(c(200, 10, 210, 10 ,210 ,20 ,200 ,20 ,200, 10),
                  ncol=2,
                  byrow=TRUE) %>%
  list() %>%
  st_polygon() %>% 
  st_sfc(crs = "+proj=longlat +datum=WGS84 +no_defs")

HS_disp <- data.frame(country = "HS",
                      PNA = F) %>% 
  left_join(ba_disp_ps, by = "country") %>% 
  st_set_geometry(HS_disp_poly)

# Los colores del mapa deben estar dados por el Post:Country
eez %>% 
  left_join(ba_disp_ps, by = c("ISO_Ter1" = "country")) %>%
  filter(!is.na(change)) %>% 
  select(country = ISO_Ter1, change, PNA) %>% 
  rbind(pipa_map_ba_disp) %>% 
  rbind(HS_disp) %>% 
  ggplot() +
  geom_sf(aes(fill = change, linetype = !PNA), color = "black") +
  ggtheme_plot() +
  scale_fill_gradientn(colors = colorRamps::matlab.like(10))
```

## Displacement

```{r, fig.cap = "Spatial representation of the mean change in the monthly allocation of fishing effort.", fig.height = 4}
eez %>% 
  left_join(ba_disp_ps, by = c("ISO_Ter1" = "country")) %>%
  filter(!is.na(change)) %>% 
  select(country = ISO_Ter1, change, PNA) %>% 
  rbind(pipa_map_ba_disp) %>% 
  rbind(HS_disp) %>% 
  filter(!country %in% c("HS", "PNG")) %>% 
  ggplot() +
  geom_sf(aes(fill = change, linetype = !PNA), color = "black") +
  ggtheme_plot() +
  scale_fill_gradientn(colors = colorRamps::matlab.like(10))
```

## Recap

- Treated vessels fish less post-implementation
  - Only significant for purse seiners

- EEZs receive proportionally more fishing effort than before
  - Proportional change increases with proximity to PIPA

## Future work

Change in behavior

- Distance traveled
- non-fishing at-sea hours
- proportion of fishing / searching

Spatial redistribution

- Proportion is bounded, might try a binomial GLM
- Measure of "crowdness"

Concerns

- Spillover effects / treatment affecting control
